---
title: "Eufalconimorph_project"
author: "Lucas Legendre"
date: "`r Sys.Date()`"
output:  html_document:
     toc: true
     toc_float:
      collapsed: false
      smooth_scroll: false
     df_print: paged
editor_options: 
  chunk_output_type: inline
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Compiled under R version 4.5.2 (2025-10-31)

<b>WARNING</b>: edit the working directory to your preferred folder.

This document details all analyses performed in R for the study:
Legendre, L.J., Clarke, J.A. 2026. Diverging trends in the syrinx morphology of songbirds and parrots reflect independent modifications of the ancestral eufalconimorph syrinx. *In prep.*

For more information regarding the study, datasets, and analyses, please refer to the Main Text and Supplementary Information of this paper. If you have any additional questions, feel free to email me at [lucasjlegendre\@gmail.com](mailto:lucasjlegendre@gmail.com){.email}.

All references cited in this script for packages, as well as additional details on my protocol and workflow, are available in the Material and Methods section of this paper.

### Loading packages
```{r, message = FALSE}
library(tidyverse)
library(ape)
library(nlme)
library(rr2)
library(phylolm)
library(phytools)
library(evobiR)
library(geiger)
library(windex)
library(Rphylopars)
library(RColorBrewer)
library(l1ou)
library(PhylogeneticEM)
library(readxl)
library(ggtree)
library(COINr)
library(MuMIn)
```

## Data and tree

-   Data
```{r}
data<-read.csv("Euf_data.csv", header=TRUE, sep=";")

# Rename duplicate species
species2<-data$Species
for (i in 2:nrow(data)) {
  if (data$Species[i] == data$Species[i-1]) {
    species2[i]<-paste(species2[i],"2", sep="_")
  }
}

data[,1]<-species2
rownames(data)<-data$Species
```

-   Tree
```{r, message=FALSE}
tree<-read.nexus("euf_tree.trees.nex")
tree$node.label<-NULL
tree<-force.ultrametric(tree)
plotTree(tree)
setdiff(tree$tip.label, data$Species) # perfect match

# Reorder data
data<-ReorderData(tree, data)
```

## Preliminary analyses

-   Log-conversion
```{r}
logdata<-cbind(data[,c(1:4)],log(data[,c(5:15)]))
```

- Impute missing values with `Rphylopars` (Goolsby et al., 2017)
```{r, warning=FALSE}
# Data
parsdata<-logdata[,c(1,5,6,8,10)]
colnames(parsdata)[1]<-"species"

# Imputation
modelpars<-c("BM", "OU", "lambda", "EB", "star") # Evolutionary models
PPE<-list(); BICdat<-data.frame()
for (i in 1:length(modelpars)) {
  PPE[[i]]<-phylopars(trait_data = parsdata, tree = tree, model=modelpars[i])
  BICdat[i,1]<-BIC(PPE[[i]]) # BIC to compare models
}

rownames(BICdat)<-modelpars; colnames(BICdat)<-"BIC"
BICdat # The lambda model is the best fit
```


-   Extract estimates and add them to the dataset
(Note: no estimates for measurements of intrinsic muscles – 'NA' in this case refers to non-applicable, rather than missing, values)
```{r}
Cdata<-cbind(logdata[,1:4], PPE[[3]]$anc_recon[1:36,1:2], logdata[,7],
            PPE[[3]]$anc_recon[1:36,3], logdata[,9],
            PPE[[3]]$anc_recon[1:36,4], logdata[,11:15])
colnames(Cdata)<-colnames(logdata)
logdata<-Cdata
logdata[24,10]<-NA # Remove the imputed value of ST CSA for O. chimborazo (hummingbirds don't have an ST)

# Add tracheal elongation ratio (TER) as an extra trait
TER<-log(exp(logdata$Tracheal_length)/(exp(logdata$Bronchial_length)+exp(logdata$Tracheal_length)))
logdata<-cbind(logdata[,1:14], TER, logdata[,15]); colnames(logdata)[16]<-"Bodymass"
```

-   Body mass ratios for all traits
```{r}
massCdata<-data
for (i in c(5:15)) {
  massCdata[,i]<-log1p(data[,i]/data$Bodymass)
} # log1p to avoid negative values

TERm<-log1p(expm1(logdata$Tracheal_length)/(expm1(logdata$Bronchial_length)+expm1(logdata$Tracheal_length))/expm1(logdata$Bodymass)) # Add TER
massCdata<-cbind(massCdata[,1:14], TERm, massCdata[,15])
massCdata<-massCdata[,-16] # Remove body mass

# Impute missing values
# Data
Mparsdata<-massCdata[,c(1,5,6,8,10)]
colnames(Mparsdata)[1]<-"species"

# Imputation
modelpars<-c("BM", "OU", "lambda", "EB", "star") # Evolutionary models
PPEM<-list(); BICdatM<-data.frame()
for (i in 1:length(modelpars)) {
  PPEM[[i]]<-phylopars(trait_data = Mparsdata, tree = tree, model=modelpars[i])
  BICdatM[i,1]<-BIC(PPEM[[i]]) # BIC to compare models
}
rownames(BICdatM)<-modelpars; colnames(BICdatM)<-"BIC"
BICdatM # The lambda model is the best fit

Cmdata<-cbind(massCdata[,1:4], PPEM[[3]]$anc_recon[1:36,1:2], massCdata[,7],
            PPEM[[3]]$anc_recon[1:36,3], massCdata[,9],
            PPEM[[3]]$anc_recon[1:36,4], massCdata[,11:15])
colnames(Cmdata)<-c(colnames(massCdata)[1:15])
massCdata<-Cmdata
massCdata[24,10]<-NA
```

-   Phenograms (for all traits)
```{r}
# For traits
for (i in 5:ncol(logdata)) {
  trait<-logdata[,i]; names(trait)<-rownames(logdata)
  trait<-na.omit(trait)
  treena<-drop.tip(tree, setdiff(tree$tip.label, names(trait)))
  phenogram(treena, trait, ,ftype="i", spread.labels=TRUE,
    spread.cost=c(1,0),fsize=0.7,color=palette()[4],
    xlab="Time (Ma)",ylab=paste("log",colnames(logdata)[i]),cex.axis=0.8,
    axes=FALSE,las=1, lwd=1)
  axis(1,at=round(seq(0,max(nodeHeights(treena)),length.out=6),1),
    label=round(seq(max(nodeHeights(treena)),0,length.out=6),1),
    cex.axis=0.8)
  axis(2,las=1,cex.axis=0.8)
  grid()
  title(paste(colnames(logdata[i])))
}

# For ratios
for (i in 5:ncol(massCdata)) {
  trait<-massCdata[,i]; names(trait)<-rownames(massCdata)
  trait<-na.omit(trait)
  treena<-drop.tip(tree, setdiff(tree$tip.label, names(trait)))
  phenogram(treena, trait, ,ftype="i", spread.labels=TRUE,
    fsize=0.7,color=palette()[4],
    xlab="Time (Ma)",ylab=paste("log",colnames(massCdata)[i],"/ body mass"),cex.axis=0.8,
    axes=FALSE,las=1, lwd=1)
  axis(1,at=round(seq(0,max(nodeHeights(treena)),length.out=6),1),
    label=round(seq(max(nodeHeights(treena)),0,length.out=6),1),
    cex.axis=0.8)
  axis(2,las=1,cex.axis=0.8)
  grid()
  title(paste(colnames(massCdata[i]), "/ body mass"))
}
```

-   Evolutionary model (for all traits) in `geiger` (Pennell et al., 2014)
```{r, warning=FALSE}
### Phylogenetic signal (Pagel's lambda)
var=list(); phy=list(); phydat=data.frame()
for (i in 5:16) {
  var<-logdata[,i]; names(var)<-rownames(logdata)
  var2<-var[!is.na(var)]
  treevar<-drop.tip(tree,setdiff(tree$tip.label,names(var2)))
  phy[[i]]<-phylosig(treevar, var2, method="lambda", test=TRUE)
  phydat[i,1]<-colnames(logdata)[i]; phydat[i,2]<-phy[[i]]$lambda; phydat[i,3]<-phy[[i]]$P
}
colnames(phydat)<-c("Variable", "Lambda", "P")
phydat[c(5:16),]

### Checking evolutionary model with fitContinuous
models=c("BM", "OU", "EB", "rate_trend","lambda", "white") # (For more info, check '?fitContinuous')

var=list(); fit=list(); mod=list()
for (i in 5:16) {
  var<-logdata[,i]; names(var)<-rownames(logdata)
  var2<-var[!is.na(var)]
  treevar<-drop.tip(tree,setdiff(tree$tip.label,names(var2)))
  for (m in 1:length(models)) {
    fit[[m]]=fitContinuous(treevar, var2, model=models[m], ncores=2)
  }
  mod[[i]]<-modSel.geiger(fit[[1]],fit[[2]],fit[[3]],fit[[4]],fit[[5]],fit[[6]])
}
mod
```
High phylogenetic signal (lambda ≥ 0.75) for all variables. Phylogenetic signal is significant (p < 0.05) for all of them except SS CSA. 
 
Best evolutionary model is: 
- a lambda model for bronchial length, TL CSA, ST CSA, labia length, LL CSA, and TER;
- a Brownian Motion (BM) model for tracheal length, tracheal diameter, SP CSA, and body mass;
- an Ornstein-Uhlenbeck (OU) model for bronchial diameter and SS CSA;

-   Same with ratios
```{r, warning=FALSE}
var=list(); phy=list(); phydat=data.frame()
for (i in 5:15) {
  var<-massCdata[,i]; names(var)<-rownames(massCdata)
  var2<-var[!is.na(var)]
  treevar<-drop.tip(tree,setdiff(tree$tip.label,names(var2)))
  phy[[i]]<-phylosig(treevar, var2, method="lambda", test=TRUE)
  phydat[i,1]<-paste(colnames(massCdata)[i], "/ body mass"); phydat[i,2]<-phy[[i]]$lambda; phydat[i,3]<-phy[[i]]$P
}
colnames(phydat)<-c("Variable", "Lambda", "P")
phydat[c(5:15),]

### Checking evolutionary model with fitContinuous
models=c("BM", "OU", "EB", "rate_trend","lambda", "white") # (For more info, check '?fitContinuous')

var=list(); fit=list(); mod=list()
for (i in 5:15) {
  var<-massCdata[,i]; names(var)<-rownames(massCdata)
  var2<-var[!is.na(var)]
  treevar<-drop.tip(tree,setdiff(tree$tip.label,names(var2)))
  for (m in 1:length(models)) {
    fit[[m]]=fitContinuous(treevar, var2, model=models[m], ncores=2)
  }
  mod[[i]]<-modSel.geiger(fit[[1]],fit[[2]],fit[[3]],fit[[4]],fit[[5]],fit[[6]])
}
mod
```
Phylogenetic signal is high (lambda ≥ 0.75), except for LL CSA (lambda = 0.64). It is significant (p < 0.05) for all ratios. 
Best evolutionary model is: 
- a lambda model for ratios of bronchial length, TL CSA, ST CSA, labia length, and LL CSA; 
- a BM model for ratios of tracheal length, tracheal diameter, and bronchial diameter; 
- an OU model for ratios of SS CSA, SP CSA, and TER.

-   Ancestral State Reconstruction (using Brownian Motion – a good model for all traits except SS CSA and bronchial diameter) in `phytools` (Revell, 2024)
```{r}
pal<-c("yellow1","orange1","orangered3","purple4")
dataplot=list(); fit=list(); obj=list()
for (i in c(5:16)) {
  dataplot[[i]]<-logdata[,i]; names(dataplot[[i]])<-rownames(logdata)
  dataplot[[i]]<-na.omit(dataplot[[i]])
  treeplot<-drop.tip(tree, setdiff(tree$tip.label, names(dataplot[[i]])))
  fit[[i]]<-fastAnc(treeplot, dataplot[[i]], vars=TRUE, CI=TRUE)
  obj[[i]]<-setMap(contMap(treeplot, dataplot[[i]], plot=FALSE),
                   colors=pal)
  plot(obj[[i]])
  title(paste('Ancestral state reconstruction for', colnames(logdata)[i]))
}
```
Possible shifts in Passeriformes (tracheal length and diameter, bronchial diameter, TL CSA, ST CSA, SS CSA), in Psittacopasseres (bronchial length), in Passeri (SP CSA), in Psittaciformes (LL CSA), in both Psittacopasseres and Passeriformes (body mass), and in both Passeriformes and Falconinae (labia length). TER does not show any conspicuous pattern.

-   Ancestral State Reconstruction for ratios (using Brownian Motion – a good model for all traits except SS CSA, SP CSA, and TER)
```{r}
dataplot=list(); fit=list(); obj=list()
for (i in c(5:15)) {
  dataplot[[i]]<-massCdata[,i]; names(dataplot[[i]])<-rownames(massCdata)
  dataplot[[i]]<-na.omit(dataplot[[i]])
  treeplot<-drop.tip(tree, setdiff(tree$tip.label, names(dataplot[[i]])))
  fit[[i]]<-fastAnc(treeplot, dataplot[[i]], vars=TRUE, CI=TRUE)
  obj[[i]]<-setMap(contMap(treeplot, dataplot[[i]], plot=FALSE),
                   colors=pal)
  plot(obj[[i]])
  title(paste('Ancestral state reconstruction for', colnames(massCdata)[i], "/ body mass"))
}
```
Less clear than for individual traits; possibles shifts in Passeriformes (tracheal and bronchial length, tracheal and bronchial diameter, TER), Psittacopasseres (TL CSA), both Psittaciformes and Passeriformes (LL CSA), but the pattern is less conspicuous for all ratios. **One shift seems to be very conspicuous**: a higher ratio of labia length for Passeriformes.

-   Boxplots by order for all traits
```{r}
# For individual traits
## Subset eufalconimorphs
eufalc<-logdata %>% filter(Order=="Passeriformes"|
                             Order=="Psittaciformes"|
                             Order=="Falconiformes")
eufalctree<-drop.tip(tree, setdiff(tree$tip.label, eufalc$Species))

## Plot
for (i in 5:16) {
  print(ggplot(eufalc, aes(x=Order, y=eufalc[,i], fill=Order)) +
  geom_boxplot() +
  #geom_jitter(shape=16, position=position_jitter(0.2)) +
  scale_fill_brewer(palette="Accent") +
  ylab(colnames(eufalc)[i]) +
  ggtitle(paste("Distribution of", colnames(eufalc)[i], "per order")))
}

# For ratios
## Subset eufalconimorphs
eufalcmass<-massCdata %>% filter(Order=="Passeriformes"|
                             Order=="Psittaciformes"|
                             Order=="Falconiformes")

for (i in 5:15) {
  print(ggplot(eufalcmass, aes(x=Order, y=eufalcmass[,i], fill=Order)) +
  geom_boxplot() +
  #geom_jitter(shape=16, position=position_jitter(0.2)) +
  scale_fill_brewer(palette="Accent") +
  ylab(paste(colnames(eufalcmass)[i], "/ body mass")) +
  ggtitle(paste("Distribution of", colnames(eufalcmass)[i], "/ body mass per order")))
}
```
There seems to be some differences – let's test for it...

- Phylogenetic ANOVA for three orders within Eufalconimorpha
```{r, warning=FALSE}
# For individual traits
datanova=list(); treenova=list(); phylanov=list()
for (i in 5:16) {
  datanova[[i]]<-eufalc[,i]; names(datanova[[i]])<-rownames(eufalc)
  datanova[[i]]<-na.omit(datanova[[i]])
  treenova[[i]]<-drop.tip(tree, setdiff(tree$tip.label, names(datanova[[i]])))
  phylanov[[i]]<-phylANOVA(treenova[[i]], as.factor(eufalc$Order), datanova[[i]], p.adj="BH")
}
phylanov

# For ratios
datanovamass=list(); treenovamass=list(); phylanovmass=list()
for (i in 5:15) {
  datanovamass[[i]]<-eufalcmass[,i]; names(datanovamass[[i]])<-rownames(eufalcmass)
  datanovamass[[i]]<-na.omit(datanovamass[[i]])
  treenovamass[[i]]<-drop.tip(tree, setdiff(tree$tip.label, names(datanovamass[[i]])))
  phylanovmass[[i]]<-phylANOVA(treenovamass[[i]], eufalcmass$Order, datanovamass[[i]],
                               p.adj="BH")
}
phylanovmass
```
None of the variables show any significant differences between orders.

## Test for shifts in syringeal traits

Using an OU model in `l1ou` (Khabbazian et al., 2016; see also Zhang et al., 2024).

-   Analyses performed for all traits at once
```{r, message=FALSE}
eModel=list()
for (i in c(5:16)) {
  trait<-logdata[,i]; names(trait)<-rownames(logdata)
  trait<-na.omit(trait)
  treena<-drop.tip(tree, setdiff(tree$tip.label, names(trait)))
  OUdat<-adjust_data(treena, trait)
  eModel[[i]]<-estimate_shift_configuration(OUdat$tree, OUdat$Y, 
                                            criterion="pBIC",
                                            root.model="OUfixedRoot")
  nEdges <- Nedge(treena)
  ew <- rep(1,nEdges)
  ew[eModel[[i]]$shift.configuration] <- 3
  plot(eModel[[i]], cex=0.5, label.offset=0.02, edge.width=ew,
       main=paste("Shift scores for", colnames(logdata[i]), sep=" "))
}

## Print list of shifts for each model and plot the shifts on the tree
par(mfrow=c(1,1))
for (i in c(5:16)) {
      model.profile<-profile(eModel[[i]])
      plot(model.profile$nShifts, model.profile$scores, 
           main=paste("Shift scores for", colnames(logdata[i]), sep=" "))
      print(paste(colnames(logdata)[i], ": ",
                  eModel[[i]]$shift.configuration))
}
```
We have significant shifts for bronchial length, TL CSA, ST CSA, labia length, LL CSA, SP CSA, and TER. Among these, the shifts for SP CSA and TER are *autapomorphic*.

Other shifts indicate a trend for major clades: 
- Bronchial length: **significant decrease in Psittacopasseres**
- TL CSA: **significant decrease in Passeriformes** 
- ST CSA: **significant decrease in Passeriformes** 
- Labia length: **significant decrease in Neoaves**, with a **subsequent increase in Pelecaniformes** (*Nannopterum auritum* + *Ardea herodias*) and a **subsequent further decrease in Arinae** (*Psittacara*, *Amazona*, and *Myiopsitta*)
- LL CSA: **significant decrease in Neoaves** and **subsequent increase in Pelecaniformes** (*Nannopterum auritum* + *Ardea herodias*)

- Same with ratios
```{r, message=FALSE}
eModel=list()
for (i in c(5:15)) {
  trait<-massCdata[,i]; names(trait)<-rownames(massCdata)
  trait<-na.omit(trait)
  treena<-drop.tip(tree, setdiff(tree$tip.label, names(trait)))
  OUdat<-adjust_data(treena, trait)
  eModel[[i]]<-estimate_shift_configuration(OUdat$tree, OUdat$Y, 
                                            criterion="pBIC",
                                            root.model="OUfixedRoot")
  nEdges <- Nedge(treena)
  ew <- rep(1,nEdges)
  ew[eModel[[i]]$shift.configuration] <- 3
  plot(eModel[[i]], cex=0.5, label.offset=0.02, edge.width=ew,
       main=paste("Shift scores for",colnames(logdata[i]), "/ Body mass", sep=" "))
}

## Print list of shifts for each model and plot the shifts on the tree
par(mfrow=c(1,1))
for (i in c(5:15)) {
      model.profile<-profile(eModel[[i]])
      plot(model.profile$nShifts, model.profile$scores,
           main=paste("Shift scores for", colnames(massCdata[i]), sep=" "))
      print(paste(colnames(massCdata)[i], ": ", eModel[[i]]$shift.configuration))
}
```
We have significant shifts for ratios of TL CSA, ST CSA, labia length, LL CSA, SS CSA, and TER. Among these, the shifts for ratios of TL CSA and SS CSA are *autapomorphic*.

Other shifts indicate a trend for major clades:
- Ratio of ST CSA / body mass: **significant increase in Neoaves** and **subsequent decrease in Gruimorphae**
- Ratio of labia length / body mass: **significant increase in Passeriformes**
- Ratio of LL CSA / body mass: **significant increase in Passeriformes**
- Ratio of TER / body mass: **significant increase in (Passeroidea + Muscicapoidea)**

We can study this further by exploring multivariate OU shifts with package `PhylogeneticEM` (Bastide et al., 2017, 2018).

- Dimensionality reduction, using phylogenetic PCA in `phytools`
```{r}
dataOU<-logdata[-24,-c(1:4,13:14)] # Remove specimen info and traits with missing values

PCAdat<-phyl.pca(tree, dataOU)

# Proportion of variance explained by each PC
PCAvar<-as.data.frame(diag(PCAdat$Eval)/sum(PCAdat$Eval)*100)
PCAvar<-cbind(rownames(PCAvar), PCAvar); colnames(PCAvar)<-c("PC", "value")
PCAvar
PCAvar$PC<-factor(PCAvar$PC, levels=rownames(PCAvar)) # Correct order for the PCs
ggplot(PCAvar, aes(PC, value)) + geom_bar(stat="identity")
PCAfull<-as.data.frame(PCAdat$S)
ggplot(PCAfull, aes(PC1, PC2)) + 
  geom_point() +
  geom_text(label=rownames(PCAfull),
    nudge_x = 0.25, nudge_y = 0.25, 
    check_overlap = T)

# Check correlation between body mass and PC1
PCs<-PCAfull[,c(1,2)]
BMlog<-as.data.frame(logdata$Bodymass); rownames(BMlog)<-rownames(logdata)
BMlog<-BMlog[-24,,FALSE] # Remove extra row for hummingbird
BMPC<-cbind(PCs, BMlog); colnames(BMPC)<-c("PC1", "PC2", "Body_mass")
treePC<-drop.tip(tree, setdiff(tree$tip.label, rownames(BMPC)))
## PGLS with PC1
BMPC1<-gls(PC1~Body_mass, BMPC, 
               correlation=corPagel(1, phy=treePC, form=~1), method="ML")
summary(BMPC1); R2_pred(BMPC1)
ggplot(BMPC, aes(Body_mass, PC1)) +
  geom_point(size=5) +
  geom_text(aes(label=rownames(BMPC)),hjust=-0.1, vjust=0.4) +
  xlab("ln body mass (g)") +
  ylab("PC1") +
  geom_abline(intercept=BMPC1$coefficients[1], slope=BMPC1$coefficients[2],
              linewidth=1.3) +
  theme(axis.text=element_text(size=15),
        axis.title=element_text(size=15),
        legend.text=element_text(size=15),
        legend.title=element_text(size=17))
## PGLS with PC2
BMPC2<-gls(PC2~Body_mass, BMPC, 
               correlation=corPagel(1, phy=treePC, form=~1), method="ML")
summary(BMPC2); R2_pred(BMPC2)
ggplot(BMPC, aes(Body_mass, PC2)) +
  geom_point(size=5) +
  geom_text(aes(label=rownames(BMPC)),hjust=-0.1, vjust=0.4) +
  xlab("ln body mass (g)") +
  ylab("PC2") +
  geom_abline(intercept=BMPC2$coefficients[1], slope=BMPC2$coefficients[2],
              linewidth=1.3) +
  theme(axis.text=element_text(size=15),
        axis.title=element_text(size=15),
        legend.text=element_text(size=15),
        legend.title=element_text(size=17))
```
The first two PCs explain 87% of the variance in the original dataset. PC1 is highly correlated to body mass.

- Estimate the shifts in the PCs and visualize them on the tree
```{r, message =FALSE}
PCAdatviz<-as.data.frame(t(PCAdat$S[,c(1:2)]))

max(abs(sqrt(nrow(logdata))), 10) # Maximum number of shifts (here, 10)

treePCA<-drop.tip(tree, setdiff(tree$tip.label, rownames(dataOU)))

res <- PhyloEM(phylo = treePCA,
               Y_data = PCAdatviz,
               process = "scOU",   
               random.root = TRUE,
               stationary.root = TRUE,
               nbr_alpha = 100,
               K_max = 10,
               parallel_alpha = TRUE,
               Ncores = 4)
res

# Plot
plot(res) # Shifts on the tree
plot_criterion(res) # model selection criterion: one shift is the best model
```
**One shift** for Passeriformes, similar to what is obtained with TL CSA and ST CSA, and with mass-corrected labia length and LL CSA (also see ASR above).

This is an interesting result, but how does it vary (or not) if we include traits for intrinsic muscles? We did not include them due to NAs for non-Psittacopasseres and *Contopus*.

Let's rerun the analysis with all traits, but only for songbirds and parrots:

- Phylogenetic PCA
```{r}
dataOUall<-na.omit(logdata[,-c(1:4)]) # Remove specimen info
treeIM<-drop.tip(tree, setdiff(tree$tip.label, rownames(dataOUall))) # Prune the tree

PCAdatall<-phyl.pca(treeIM, dataOUall)

# Proportion of variance explained by each PC
PCAvarall<-as.data.frame(diag(PCAdatall$Eval)/sum(PCAdatall$Eval)*100)
PCAvarall<-cbind(rownames(PCAvarall), PCAvarall); colnames(PCAvarall)<-c("PC", "value")
PCAvarall
PCAvarall$PC<-factor(PCAvarall$PC, levels=rownames(PCAvarall)) # Correct order for the PCs
ggplot(PCAvarall, aes(PC, value)) + geom_bar(stat="identity")
```
The first two PCs explain ~82% of the variance in the original dataset.

- Estimate the shifts in the PCs and visualize them on the tree
```{r}
PCAdatallviz<-as.data.frame(t(PCAdatall$S[,c(1:2)]))

max(abs(sqrt(nrow(dataOUall))), 10) # Maximum number of shifts (here, 10)

resall <- PhyloEM(phylo = treeIM,
               Y_data = PCAdatallviz,
               process = "scOU",   
               random.root = TRUE,
               stationary.root = TRUE,
               nbr_alpha = 100,
               K_max = 10,
               parallel_alpha = TRUE,
               Ncores = 4)
resall

# Plot
plot(resall) # Shifts on the tree
plot_criterion(resall) # model selection criterion: no shifts is the best model
```
The lack of extra taxa blurs the signal, but it could also be that the IM traits have very little signal as well – a larger sample size for Psittacopasseres might help.

Since there seems to be a major shift in Passeriformes for both labia traits (length and CSA) that this might be related to body mass (significant shifts for ratios), we should investigate the relationship between those traits and body mass a bit further using phylogenetic regressions.

### Relationship between labia traits and body mass using PGLS

Phylogenetic generalized least squares in `phylolm` (Ho & Ané, 2014).

#### For (labia length ~ body mass)

```{r}
model<-c("BM", "OUrandomRoot", "OUfixedRoot", "lambda", "kappa", "delta", "EB")
LLmodel<-list()
for (i in 1:length(model)) {
  LLmodel[[i]]<-phylolm(LL~Bodymass, logdata, tree, model=paste(model[[i]]))
}
mod2<-modSel.phylolm(LLmodel[[1]], LLmodel[[2]], LLmodel[[3]], LLmodel[[4]],
                     LLmodel[[5]], LLmodel[[6]], LLmodel[[7]], tree=tree,
                     method="phylolm")
mod2 # The lambda model shows the best fit

lambdaLL<-LLmodel[[4]]; summary(lambdaLL) 
```
Model is highly significant (p < 0.0001), but the pseudo R-squared is pretty low (0.35).

- Same with `ape` (Paradis & Schliep, 2019) and `rr2` (Ives & Li, 2018; Ives, 2019)
```{r}
lambdaLL2<-gls(LL~Bodymass, logdata, 
               correlation=corPagel(1, phy=tree, form=~1), method="ML")
summary(lambdaLL2)
R2_pred(lambdaLL2)
```
Pseudo R-squared is fairly high (0.71).

- Let's plot it and check the residuals
```{r}
# Plot
dupli<-logdata$Order; dupli[21:30]<-"Other"
logdata2<-cbind(logdata[,1], dupli, logdata[,3:16])
colnames(logdata2)[c(1,2)]<-c("Species","Order")

hull <- logdata2[-c(21:30),] %>% group_by(Order) %>% 
  slice(chull(Bodymass, LL)) # Convex hulls for eufalconimorphs

ggplot(logdata2, aes(Bodymass, LL, color=Order)) +
  geom_point(size=5) +
  geom_text(aes(label=Species),hjust=-0.1, vjust=0.4) +
  xlab("ln body mass (g)") +
  ylab("ln labia length (µm)") +
  geom_abline(intercept=lambdaLL$coefficients[1], slope=lambdaLL$coefficients[2],
              linewidth=1.3) +
  theme(axis.text=element_text(size=15),
        axis.title=element_text(size=15),
        legend.text=element_text(size=15),
        legend.title=element_text(size=17)) +
  scale_colour_brewer("Order", palette="Set1") +
  aes(fill=Order) +
  geom_polygon(data=hull, alpha=0.5)

# Normality test
shapiro.test(chol(solve(vcv(tree)))%*%lambdaLL$residuals) # non normal

# Checking the residuals
res<-residuals(lambdaLL2, type="n")
plot(lambdaLL2, resid(., type="n")~fitted(.), 
     main="Normalized Residuals v Fitted Values",abline=c(0,0))
qqnorm(res); qqline(res)

# Removing the outliers
logdata2new<-logdata2[-which(logdata$Species=="Struthio_camelus"),]
colnames(logdata2new)[1]<-"Species"
treenew<-drop.tip(tree, setdiff(tree$tip.label, logdata2new$Species))

# New PGLS
## With ape
lambdaLL2new<-gls(LL~Bodymass, logdata2new, 
                 correlation=corPagel(1, phy=treenew, form=~1), method="ML") 
summary(lambdaLL2new)
R2_pred(lambdaLL2new) # slightly lower
## With phylolm
lambdaLLnew<-phylolm(LL~Bodymass, logdata2new, treenew, model="lambda")
summary(lambdaLLnew) # Higher p-value, but still highly significant

shapiro.test(chol(solve(vcv(treenew)))%*%residuals(lambdaLLnew))
resnew<-resid(lambdaLL2new, type="n")
plot(lambdaLL2new, resid(., type="n")~fitted(.), main="Normalized Residuals v Fitted Values",abline=c(0,0)) # Looks better
qqnorm(resnew); qqline(resnew) # Looks better

# Plot
ggplot(logdata2new, aes(Bodymass, LL, color=Order)) +
  geom_point(size=5) +
  # geom_text(aes(label=Species),hjust=-0.1, vjust=0.4) +
  xlab("ln body mass (g)") +
  ylab("ln labia length (µm)") +
  geom_abline(intercept=lambdaLL2new$coefficients[1],
              slope=lambdaLL2new$coefficients[2], linewidth=1.3) +
  theme(axis.text=element_text(size=15),
        axis.title=element_text(size=15),
        legend.text=element_text(size=15),
        legend.title=element_text(size=17)) +
  scale_colour_brewer("Order", palette="Set1") +
  aes(fill=Order) +
  geom_polygon(data=hull, alpha=0.5)
```

#### For (LL CSA ~ body mass)

```{r}
model<-c("BM", "OUrandomRoot", "OUfixedRoot", "lambda", "kappa", "delta", "EB")
LLCSAmodel<-list()
for (i in 1:length(model)) {
  LLCSAmodel[[i]]<-phylolm(LL_CSA~Bodymass, logdata, tree, model=paste(model[[i]]))
}
mod2<-modSel.phylolm(LLCSAmodel[[1]], LLCSAmodel[[2]], LLCSAmodel[[3]],
                     LLCSAmodel[[4]], LLCSAmodel[[5]], LLCSAmodel[[6]],
                     LLCSAmodel[[7]], tree=tree, method="phylolm")
mod2 # The lambda model shows the best fit

lambdaLLCSA<-LLCSAmodel[[4]]; summary(lambdaLLCSA) 
```
Model is highly significant (p < 0.0001), but the pseudo R-squared is fairly low (0.58).

- Same with `ape` (Paradis & Schliep, 2019) and `rr2` (Ives & Li, 2018; Ives, 2019)
```{r}
lambdaLLCSA2<-gls(LL_CSA~Bodymass, logdata, 
               correlation=corPagel(1, phy=tree, form=~1), method="ML")
summary(lambdaLLCSA2)
R2_pred(lambdaLLCSA2)
```
Pseudo R-squared is fairly high (0.75).

- Let's plot it and check the residuals
```{r}
# Plot
hull2 <- logdata2[-c(21:30),] %>% group_by(Order) %>% 
  slice(chull(Bodymass, LL_CSA)) # Convex hulls for eufalconimorphs

ggplot(logdata2, aes(Bodymass, LL_CSA, color=Order)) +
  geom_point(size=5) +
  # geom_text(aes(label=Species),hjust=-0.1, vjust=0.4) +
  xlab("ln body mass (g)") +
  ylab("ln LL CSA (µm^2)") +
  geom_abline(intercept=lambdaLLCSA$coefficients[1],
              slope=lambdaLLCSA$coefficients[2], linewidth=1.3) +
  theme(axis.text=element_text(size=15),
        axis.title=element_text(size=15),
        legend.text=element_text(size=15),
        legend.title=element_text(size=17)) +
  scale_colour_brewer("Order", palette="Set1") +
  aes(fill=Order) +
  geom_polygon(data=hull2, alpha=0.5)

# Normality test
shapiro.test(chol(solve(vcv(tree)))%*%lambdaLLCSA2$residuals) # non normal

# Checking the residuals
res<-residuals(lambdaLLCSA2, type="n")
plot(lambdaLLCSA2, resid(., type="n")~fitted(.), 
     main="Normalized Residuals v Fitted Values",abline=c(0,0))
qqnorm(res); qqline(res)

# New PGLS without the ostrich
## With ape
lambdaLLCSA2new<-gls(LL_CSA~Bodymass, logdata2new, 
                 correlation=corPagel(1, phy=treenew, form=~1), method="ML") 
summary(lambdaLLCSA2new)
R2_pred(lambdaLLCSA2new) # Slightly lower, but still ok
## With phylolm
lambdaLLCSAnew<-phylolm(LL_CSA~Bodymass, logdata2new, treenew, model="lambda")
summary(lambdaLLCSAnew) # Higher p-value, but still highly significant

shapiro.test(chol(solve(vcv(treenew)))%*%residuals(lambdaLLCSAnew)) # Still not normal
resnew<-resid(lambdaLLCSA2new, type="n")
plot(lambdaLLCSA2new, resid(., type="n")~fitted(.), main="Normalized Residuals v Fitted Values",abline=c(0,0)) # Looks better
qqnorm(resnew); qqline(resnew) # Still not ideal

# Plot
ggplot(logdata2new, aes(Bodymass, LL_CSA, color=Order)) +
  geom_point(size=5) +
  # geom_text(aes(label=Species),hjust=-0.1, vjust=0.4) +
  xlab("ln body mass (g)") +
  ylab("ln labia length (µm)") +
  geom_abline(intercept=lambdaLLCSA2new$coefficients[1],
              slope=lambdaLLCSA2new$coefficients[2], linewidth=1.3) +
  theme(axis.text=element_text(size=15),
        axis.title=element_text(size=15),
        legend.text=element_text(size=15),
        legend.title=element_text(size=17)) +
  scale_colour_brewer("Order", palette="Set1") +
  aes(fill=Order) +
  geom_polygon(data=hull2, alpha=0.5)
```

#### For (TER ~ body mass)

```{r}
model<-c("BM", "OUrandomRoot", "OUfixedRoot", "lambda", "kappa", "delta", "EB")
TERmodel<-list()
for (i in 1:length(model)) {
  TERmodel[[i]]<-phylolm(TER~Bodymass, logdata, tree, model=paste(model[[i]]))
}
mod2<-modSel.phylolm(TERmodel[[1]], TERmodel[[2]], TERmodel[[3]], TERmodel[[4]], TERmodel[[5]], 
               TERmodel[[6]], TERmodel[[7]], tree=tree, method="phylolm")
mod2 # The lambda model shows the best fit

lambdaTER<-TERmodel[[4]]; summary(lambdaTER)
```
The model is significant (p = 0.02), but the pseudo R-squared is very low (0.13).

- Same with `ape` (Paradis & Schliep, 2019) and `rr2` (Ives & Li, 2018; Ives, 2019)
```{r}
lambdaTER2<-gls(TER~Bodymass, logdata, 
               correlation=corPagel(0.5, phy=tree, form=~1), method="ML")
summary(lambdaTER2)
R2_pred(lambdaTER2)
```
Still very low (0.3)!

- Let's plot it and check the residuals
```{r}
# Plot
hull3 <- logdata2[-c(21:30),] %>% group_by(Order) %>% 
  slice(chull(Bodymass, TER)) # Convex hulls for eufalconimorphs

ggplot(logdata2, aes(Bodymass, TER, color=Order)) +
  geom_point(size=5) +
  # geom_text(aes(label=Species),hjust=-0.1, vjust=0.4) +
  xlab("ln body mass (g)") +
  ylab("ln TER") +
  geom_abline(intercept=lambdaTER2$coefficients[1],
              slope=lambdaTER2$coefficients[2], linewidth=1.3) +
  theme(axis.text=element_text(size=15),
        axis.title=element_text(size=15),
        legend.text=element_text(size=15),
        legend.title=element_text(size=17)) +
  scale_colour_brewer("Order", palette="Set1") +
  aes(fill=Order) +
  geom_polygon(data=hull3, alpha=0.5)

# Normality test
shapiro.test(chol(solve(vcv(tree)))%*%lambdaTER2$residuals) # non normal

# Checking the residuals
res<-residuals(lambdaTER2, type="n")
plot(lambdaTER2, resid(., type="n")~fitted(.), 
     main="Normalized Residuals v Fitted Values",abline=c(0,0))
qqnorm(res); qqline(res)

# New PGLS without S. camelus, O. chimborazo, A. viridigenalis, and P. holochlorus 2 (all outliers!)
logdata3<-logdata2[-c(14,24,30,35),]
tree3<-drop.tip(tree, setdiff(tree$tip.label, logdata3$Species))
## With ape
lambdaTER2new<-gls(TER~Bodymass, logdata3, 
                 correlation=corPagel(0.3, phy=tree3, form=~1), method="ML") 
summary(lambdaTER2new)
R2_pred(lambdaTER2new) # Negative: terrible signal! Not structured phylogenetically
## With phylolm
lambdaTERnew<-phylolm(TER~Bodymass, logdata3, tree3, model="lambda")
summary(lambdaTERnew) # Not significant, very low pseudo R2 – as expected

shapiro.test(chol(solve(vcv(tree3)))%*%residuals(lambdaTERnew)) # Still not normal
resnew<-resid(lambdaTER2new, type="n")
plot(lambdaTER2new, resid(., type="n")~fitted(.), main="Normalized Residuals v Fitted Values",abline=c(0,0)) # Looks better
qqnorm(resnew); qqline(resnew) # Still not ideal

# Plot
hull4 <- logdata3[-c(20:27),] %>% group_by(Order) %>% 
  slice(chull(Bodymass, TER)) # New convex hulls for eufalconimorphs

ggplot(logdata3, aes(Bodymass, TER, color=Order)) +
  geom_point(size=5) +
  # geom_text(aes(label=Species),hjust=-0.1, vjust=0.4) +
  xlab("ln body mass (g)") +
  ylab("ln TER") +
  geom_abline(intercept=lambdaTERnew$coefficients[1],
              slope=lambdaTERnew$coefficients[2], linewidth=1.3) +
  theme(axis.text=element_text(size=15),
        axis.title=element_text(size=15),
        legend.text=element_text(size=15),
        legend.title=element_text(size=17)) +
  scale_colour_brewer("Order", palette="Set1") +
  aes(fill=Order) +
  geom_polygon(data=hull4, alpha=0.5)
```

#### For (ST CSA ~ body mass)

```{r}
model<-c("BM", "OUrandomRoot", "OUfixedRoot", "lambda", "kappa", "delta", "EB")

logdataST<-logdata[-24,]; treeST<-drop.tip(tree, setdiff(tree$tip.label, logdataST$Species))
STCSAmodel<-list()
for (i in 1:length(model)) {
  STCSAmodel[[i]]<-phylolm(ST_CSA~Bodymass, logdataST, treeST, model=paste(model[[i]]))
}
mod2<-modSel.phylolm(STCSAmodel[[1]], STCSAmodel[[2]], STCSAmodel[[3]], STCSAmodel[[4]], STCSAmodel[[5]], 
               STCSAmodel[[6]], STCSAmodel[[7]], tree=tree, method="phylolm")
mod2 # The kappa model shows the best fit

OUrSTCSA<-STCSAmodel[[5]]; summary(OUrSTCSA)
```
The model is highly significant (p < 0.001) and the pseudo R-squared is high (0.71)!

- Same with `ape` (Paradis & Schliep, 2019) and `rr2` (Ives & Li, 2018; Ives, 2019), to compare
```{r}
OUrSTCSA2<-gls(ST_CSA~Bodymass, logdataST, 
               correlation=corMartins(0.1, phy=treeST, form=~1), method="ML")
summary(OUrSTCSA2)
R2_pred(OUrSTCSA2)
```
Even higher (0.8).

- Let's plot it and check the residuals
```{r}
# Plot
hull5 <- logdataST[-c(21:29),] %>% group_by(Order) %>% 
  slice(chull(Bodymass, ST_CSA)) # Convex hulls for eufalconimorphs

ggplot(logdata2, aes(Bodymass, ST_CSA, color=Order)) +
  geom_point(size=5) +
  # geom_text(aes(label=Species),hjust=-0.1, vjust=0.4) +
  xlab("ln body mass (g)") +
  ylab("ln ST CSA (µm^2)") +
  geom_abline(intercept=OUrSTCSA$coefficients[1],
              slope=OUrSTCSA$coefficients[2], linewidth=1.3) +
  theme(axis.text=element_text(size=15),
        axis.title=element_text(size=15),
        legend.text=element_text(size=15),
        legend.title=element_text(size=17)) +
  scale_colour_brewer("Order", palette="Set1") +
  aes(fill=Order) +
  geom_polygon(data=hull5, alpha=0.5)

# Normality test
shapiro.test(chol(solve(vcv(treeST)))%*%OUrSTCSA2$residuals) # non normal

# Checking the residuals
res<-residuals(OUrSTCSA2, type="n")
plot(OUrSTCSA2, resid(., type="n")~fitted(.), 
     main="Normalized Residuals v Fitted Values",abline=c(0,0))
qqnorm(res); qqline(res)

# New PGLS without the ostrich
logdataSTnew<-logdataST[-29,]; logdataSTnew$Order<-logdata2$Order[-c(24,30)]
treeSTnew<-drop.tip(treeST, setdiff(treeST$tip.label, logdataSTnew$Species))
## With ape
OUrSTCSA2new<-gls(ST_CSA~Bodymass, logdataSTnew, 
                 correlation=corMartins(0.1, phy=treeSTnew, form=~1), method="ML") 
summary(OUrSTCSA2new)
R2_pred(OUrSTCSA2new) # Slightly lower (0.75), but still ok
## With phylolm
OUrSTCSAnew<-phylolm(ST_CSA~Bodymass, logdataSTnew, treeSTnew, model="lambda")
summary(OUrSTCSAnew) # Higher p-value, but still highly significant

shapiro.test(chol(solve(vcv(treeSTnew)))%*%residuals(OUrSTCSAnew)) # Still not normal

resnew<-resid(OUrSTCSA2new, type="n")
plot(OUrSTCSA2new, resid(., type="n")~fitted(.), main="Normalized Residuals v Fitted Values",abline=c(0,0)) # Looks better
qqnorm(resnew); qqline(resnew) # Looks better

# Plot
ggplot(logdataSTnew, aes(Bodymass, ST_CSA, color=Order)) +
  geom_point(size=5) +
  # geom_text(aes(label=Species),hjust=-0.1, vjust=0.4) +
  xlab("ln body mass (g)") +
  ylab("ln ST CSA (µm^2)") +
  geom_abline(intercept=OUrSTCSA2new$coefficients[1],
              slope=OUrSTCSA2new$coefficients[2], linewidth=1.3) +
  theme(axis.text=element_text(size=15),
        axis.title=element_text(size=15),
        legend.text=element_text(size=15),
        legend.title=element_text(size=17)) +
  scale_colour_brewer("Order", palette="Set1") +
  aes(fill=Order) +
  geom_polygon(data=hull5, alpha=0.5)
```
Passerines are mostly on the lower end of the plot for both ST CSA and body mass; congruent with ASR and OU shifts for individual traits.

#### Same for eufalconimorphs only

- For (labia length ~ body mass)
```{r}
LLmodelE<-list()
for (i in 1:length(model)) {
  LLmodelE[[i]]<-phylolm(LL~Bodymass, eufalc, eufalctree, model=paste(model[[i]]))
}
mod3<-modSel.phylolm(LLmodelE[[1]], LLmodelE[[2]], LLmodelE[[3]], LLmodelE[[4]], LLmodelE[[5]], 
               LLmodelE[[6]], LLmodelE[[7]], tree=eufalctree, method="phylolm")
mod3 # The lambda model shows the best fit

lambdaLLE<-LLmodelE[[4]]; summary(lambdaLLE) # Significant, but low pseudo R-squared (0.23; as expected)

## Same with ape and rr2
lambdaLLE2<-gls(LL~Bodymass, eufalc, 
               correlation=corPagel(1, phy=eufalctree, form=~1), method="ML")
summary(lambdaLLE2)
R2_pred(lambdaLLE2) # Lower, but still ok (0.64)

# Plot
hull <- eufalc %>% group_by(Order) %>% 
  slice(chull(Bodymass, LL)) # Convex hulls for eufalconimorphs

ggplot(eufalc, aes(Bodymass, LL, color=Order)) +
  geom_point(size=5) +
  # geom_text(aes(label=Species),hjust=-0.1, vjust=0.4) +
  xlab("ln body mass (g)") +
  ylab("ln labia length (µm)") +
  geom_abline(intercept=lambdaLLE$coefficients[1], slope=lambdaLLE$coefficients[2],
              linewidth=1.3) +
  theme(axis.text=element_text(size=15),
        axis.title=element_text(size=15),
        legend.text=element_text(size=15),
        legend.title=element_text(size=17)) +
  scale_colour_brewer("Order", palette="Set1") +
  aes(fill=Order) +
  geom_polygon(data=hull, alpha=0.5)

# Normality test
shapiro.test(chol(solve(vcv(eufalctree)))%*%lambdaLLE$residuals) # non normal

# Checking the residuals
res<-residuals(lambdaLLE2, type="n")
plot(lambdaLLE2, resid(., type="n")~fitted(.),
     main="Normalized Residuals v Fitted Values",abline=c(0,0)) # Looks good
qqnorm(res); qqline(res) # Looks good
```

- For (LL CSA ~ body mass)
```{r}
LLCSAmodelE<-list()
for (i in 1:length(model)) {
  LLCSAmodelE[[i]]<-phylolm(LL_CSA~Bodymass, eufalc, eufalctree, model=paste(model[[i]]))
}
mod3<-modSel.phylolm(LLCSAmodelE[[1]], LLCSAmodelE[[2]], LLCSAmodelE[[3]], LLCSAmodelE[[4]], LLCSAmodelE[[5]], 
               LLCSAmodelE[[6]], LLCSAmodelE[[7]], tree=eufalctree, method="phylolm")
mod3 # The lambda model shows the best fit

lambdaLLCSAE<-LLCSAmodelE[[4]]; summary(lambdaLLCSAE) # Significant, but low pseudo R-squared (0.36; as expected)

## Same with ape and rr2
lambdaLLCSAE2<-gls(LL_CSA~Bodymass, eufalc, 
               correlation=corPagel(1, phy=eufalctree, form=~1), method="ML")
summary(lambdaLLCSAE2)
R2_pred(lambdaLLCSAE2) # Still ok (0.6)

# Plot
hull2 <- eufalc %>% group_by(Order) %>% 
  slice(chull(Bodymass, LL_CSA)) # Convex hulls for eufalconimorphs

ggplot(eufalc, aes(Bodymass, LL_CSA, color=Order)) +
  geom_point(size=5) +
  # geom_text(aes(label=Species),hjust=-0.1, vjust=0.4) +
  xlab("ln body mass (g)") +
  ylab("ln LL CSA (µm^2)") +
  geom_abline(intercept=lambdaLLCSAE$coefficients[1],
              slope=lambdaLLCSAE$coefficients[2], linewidth=1.3) +
  theme(axis.text=element_text(size=15),
        axis.title=element_text(size=15),
        legend.text=element_text(size=15),
        legend.title=element_text(size=17)) +
  scale_colour_brewer("Order", palette="Set1") +
  aes(fill=Order) +
  geom_polygon(data=hull2, alpha=0.5)

# Normality test
shapiro.test(chol(solve(vcv(eufalctree)))%*%lambdaLLCSAE2$residuals) # non normal

# Checking the residuals
res<-residuals(lambdaLLCSAE2, type="n")
plot(lambdaLLCSAE2, resid(., type="n")~fitted(.), 
     main="Normalized Residuals v Fitted Values",abline=c(0,0)) # Looks good
qqnorm(res); qqline(res) # Looks good
```

- For (TER ~ body mass)
```{r}
TERmodelE<-list()
for (i in 1:length(model)) {
  TERmodelE[[i]]<-phylolm(TER~Bodymass, eufalc, eufalctree, model=paste(model[[i]]))
}
mod3<-modSel.phylolm(TERmodelE[[1]], TERmodelE[[2]], TERmodelE[[3]], TERmodelE[[4]], TERmodelE[[5]], 
               TERmodelE[[6]], TERmodelE[[7]], tree=eufalctree, method="phylolm")
mod3 # The lambda model shows the best fit

lambdaTERE<-TERmodelE[[4]]; summary(lambdaTERE) # Not significant, low pseudo R-squared (0.04; as expected)

## Same with ape and rr2
lambdaTERE2<-gls(TER~Bodymass, eufalc, 
               correlation=corPagel(1, phy=eufalctree, form=~1), method="ML")
summary(lambdaTERE2)
R2_pred(lambdaTERE2) # Slightly higher, still very low (0.12)

# Plot
hull3 <- eufalc %>% group_by(Order) %>% 
  slice(chull(Bodymass, TER)) # Convex hulls for eufalconimorphs

ggplot(eufalc, aes(Bodymass, TER, color=Order)) +
  geom_point(size=5) +
  # geom_text(aes(label=Species),hjust=-0.1, vjust=0.4) +
  xlab("ln body mass (g)") +
  ylab("ln labia length (µm)") +
  geom_abline(intercept=lambdaTERE$coefficients[1], slope=lambdaTERE$coefficients[2],
              linewidth=1.3) +
  theme(axis.text=element_text(size=15),
        axis.title=element_text(size=15),
        legend.text=element_text(size=15),
        legend.title=element_text(size=17)) +
  scale_colour_brewer("Order", palette="Set1") +
  aes(fill=Order) +
  geom_polygon(data=hull3, alpha=0.5)

# Normality test
shapiro.test(chol(solve(vcv(eufalctree)))%*%lambdaTERE$residuals) # non normal

# Checking the residuals
res<-residuals(lambdaTERE2, type="n")
plot(lambdaTERE2, resid(., type="n")~fitted(.),
     main="Normalized Residuals v Fitted Values",abline=c(0,0)) # Some outliers
qqnorm(res); qqline(res) # Extreme values are off
```

- For (ST CSA ~ body mass)
```{r}
STCSAmodelE<-list()
for (i in 1:length(model)) {
  STCSAmodelE[[i]]<-phylolm(ST_CSA~Bodymass, eufalc, eufalctree,
                            model=paste(model[[i]]))
}
mod3<-modSel.phylolm(STCSAmodelE[[1]], STCSAmodelE[[2]], STCSAmodelE[[3]], STCSAmodelE[[4]], STCSAmodelE[[5]], 
               STCSAmodelE[[6]], STCSAmodelE[[7]], tree=eufalctree, method="phylolm")
mod3 # Both OU models show the best fit (same AICc value)

#Let's do some model averaging:
STCSAavg<-model.avg(list(STCSAmodelE[[2]], STCSAmodelE[[3]]))
summary(STCSAavg) # Highly significant, high pseudo R-squared (0.75)

## Same with ape and rr2
STCSAE2<-gls(ST_CSA~Bodymass, eufalc, 
               correlation=corMartins(0.1, phy=eufalctree, form=~1), method="ML")
summary(STCSAE2)
R2_pred(STCSAE2) # Higher (0.83)

# Plot
hull6 <- eufalc %>% group_by(Order) %>% 
  slice(chull(Bodymass, ST_CSA)) # Convex hulls for eufalconimorphs

ggplot(eufalc, aes(Bodymass, ST_CSA, color=Order)) +
  geom_point(size=5) +
  # geom_text(aes(label=Species),hjust=-0.1, vjust=0.4) +
  xlab("ln body mass (g)") +
  ylab("ln ST CSA (µm^2)") +
  geom_abline(intercept=STCSAavg$coefficients[1],
              slope=STCSAavg$coefficients[3], linewidth=1.3) +
  theme(axis.text=element_text(size=15),
        axis.title=element_text(size=15),
        legend.text=element_text(size=15),
        legend.title=element_text(size=17)) +
  scale_colour_brewer("Order", palette="Set1") +
  aes(fill=Order) +
  geom_polygon(data=hull6, alpha=0.5)

# Normality test
shapiro.test(chol(solve(vcv(eufalctree)))%*%STCSAE2$residuals) # non normal

# Checking the residuals
res<-residuals(STCSAE2, type="n")
plot(STCSAE2, resid(., type="n")~fitted(.), 
     main="Normalized Residuals v Fitted Values",abline=c(0,0)) # Looks good
qqnorm(res); qqline(res) # Not ideal, but not surprising
```

## Ancestral state reconstructions (ASR) for intrinsic muscles in Eufalconimorphs

Using data from our sample and the literature to reconstruct ASR with SIMMAP (Huelsenbeck & Bollback, 2003; Bollback, 2006; Revell, 2013) in `phytools`.

-   Data
```{r, warning=FALSE}
datasub<-read_excel("ASR_IM.xlsx")
rownames(datasub)<-datasub$Species
```

-   Tree
```{r}
treemcT<-read.nexus("summary_dated_clements.nex")

setdiff(datasub$Species, treemcT$tip.label) # Check for discrepancies between taxa names in our data set and in the tree

specmcT<-data.frame(treemcT$tip.label) # to look for synonyms directly
treemcT$tip.label[2191]<-"Tychaedon_barbata"
treemcT$tip.label[3024]<-"Paradoxornis_unicolor"
treemcT$tip.label[3025]<-"Paradoxornis_gularis"
treemcT$tip.label[3376]<-"Ptyonoprogne_obsoleta"
treemcT$tip.label[4787]<-"Ceblepyris_pectoralis"
treemcT$tip.label[6207]<-"Tyranniscus_nigrocapillus"
treemcT$tip.label[6506]<-"Chiroxiphia_galeata"
treemcT$tip.label[7171]<-"Astur_cooperii"

setdiff(datasub$Species, treemcT$tip.label) # Perfect match

treesub<-drop.tip(treemcT, setdiff(treemcT$tip.label, datasub$Species))
plotTree(treesub, fsize=0.2, lwd=1, type="fan", ftype="i", part=0.99)
```

-   ASR using stochastic character mapping (SIMMAP): first, let's use a 5-state trait for the number of pairs of intrinsic muscles (0 to 4).
```{r}
# Color palette
cols<-setNames(c("gold2", "#4DAF4AFF", "#E41A1CFF", "dodgerblue2", "#984EA3FF"), c(0, 1, 2, 3, 4))

x<-datasub$IM; names(x)<-datasub$Species; fac<-as.factor(x)

# Functions for Akaike weights
aic<-function(logL,k) 2*k-2*logL
aic.w<-function(aic){
    d.aic<-aic-min(aic)
    exp(-1/2*d.aic)/sum(exp(-1/2*d.aic))
}

# Build the models
logL<-sapply(c("ER","SYM","ARD"),
    function(model,tree,x) make.simmap(tree,x,model)$logL,
    tree=treesub,x=x)
logL

AIC<-mapply(aic,logL,c(1,3,6))
AIC
AIC.W<-aic.w(AIC)
AIC.W

nsim<-1000
Nsim<-round(nsim*AIC.W)
d<-if(sum(Nsim)>nsim) -1 else 1
nsim<-Nsim+d*sample(c(rep(1,abs(nsim-sum(Nsim))),
    rep(0,length(Nsim)-abs(nsim-sum(Nsim)))))
nsim

# Extract the weights and compile SIMMAP ASR for each model
#o1<-make.simmap(treesub,x,model="ER",nsim=nsim[1])
#o2<-make.simmap(treesub,x,model="SYM",nsim=nsim[2])
#o3<-make.simmap(treesub,x,model="ARD",nsim=nsim[3])
#treesstudy<-c(o1,o2,o3)
treesstudy1<-make.simmap(treesub,x,model="ARD",nsim=nsim[3])

objstudy1<-describe.simmap(treesstudy1)

# Plot the tree (to be executed all at once)
plot(treesstudy1[[1]], cols, type="fan", ftype="off", lwd=1, part=0.99)
obj<-get("last_plot.phylo", envir=.PlotPhyloEnv)
n<-Ntip(treesub)
par(lend=3)
for(i in 1:Ntip(treesub)){
    cc<-if(obj$xx[i]>0) 5 else -5
    th<-atan(obj$yy[i]/obj$xx[i])
    segments(obj$xx[i],obj$yy[i],
        obj$xx[i]+cc*cos(th),
        obj$yy[i]+cc*sin(th),lwd=4,
        col=cols[fac[treesub$tip.label[i]]])
}
legend("topleft",legend=levels(fac),
    pch=15,col=cols,
    pt.cex=1.5,cex=0.8,bty="n")

# Same with posterior probabilities (pp) at nodes, with larger pie charts for pp ≤ 0.95
pp<-summary(treesstudy1)$ace[1:treesub$Nnode,]
plot(treesstudy1[[1]], cols, type="fan", ftype="off", lwd=1,
     ylim=max(nodeHeights(treesub))*c(-1,1.1))
obj<-get("last_plot.phylo",envir=.PlotPhyloEnv)
n<-Ntip(treesub)
par(lend=3)
for(i in 1:Ntip(treesub)){
    cc<-if(obj$xx[i]>0) 5 else -5
    th<-atan(obj$yy[i]/obj$xx[i])
    segments(obj$xx[i],obj$yy[i],
        obj$xx[i]+cc*cos(th),
        obj$yy[i]+cc*sin(th),lwd=4,
        col=cols[fac[treesub$tip.label[i]]])
}
h<-max(nodeHeights(treesub))
ii<-order(rowSums(pp^2),decreasing=TRUE)
for(i in ii){
    plotrix::floating.pie(obj$xx[i+Ntip(treesub)],
        obj$yy[i+Ntip(treesub)],
        radius=if(max(pp[i,])<=0.95) 0.03*h else 0.01*h,
        x=pp[i,],col=cols,border="transparent")
}
legend(x=0.98*par()$usr[1],y=0.9*par()$usr[4],
    levels(fac),pch=16,col=cols,
    pt.cex=2,bty="n",cex=0.8)
```

-   Same with a binary trait (presence/absence of intrinsic muscles)
```{r}
# Color palette
cols<-setNames(c("royalblue","red3"), c(0, 1))

binaryIM<-datasub$IM; names(binaryIM)<-datasub$Species
binaryIM[which(binaryIM>=2)]<-1
x<-binaryIM; fac<-as.factor(x)

# Build the models
logL<-sapply(c("ER","SYM","ARD"),
    function(model,tree,x) make.simmap(tree,x,model)$logL,
    tree=treesub,x=x)
logL

AIC<-mapply(aic,logL,c(1,3,6))
AIC
AIC.W<-aic.w(AIC)
AIC.W

nsim<-1000
Nsim<-round(nsim*AIC.W)
d<-if(sum(Nsim)>nsim) -1 else 1
nsim<-Nsim+d*sample(c(rep(1,abs(nsim-sum(Nsim))),
    rep(0,length(Nsim)-abs(nsim-sum(Nsim)))))
nsim

# Extract the weights and compile SIMMAP ASR for each model
o1<-make.simmap(treesub,x,model="ER",nsim=nsim[1])
o2<-make.simmap(treesub,x,model="SYM",nsim=nsim[2])
o3<-make.simmap(treesub,x,model="ARD",nsim=nsim[3])
treesstudy<-c(o1,o2,o3)

objstudy<-describe.simmap(treesstudy)

# Plot the tree as a density map with posterior probabilities (pp) at nodes, with larger pie charts for pp ≤ 0.95
dMap<-setMap(densityMap(treesstudy,plot=FALSE), cols)
pp<-summary(treesstudy)$ace[1:treesub$Nnode,]
plot(dMap, type="fan", ftype="off", lwd=1, legend=FALSE, part=0.99, 
     ylim=max(nodeHeights(treesub))*c(-1,1.1))
obj<-get("last_plot.phylo",envir=.PlotPhyloEnv)
n<-Ntip(treesub)
par(lend=3)
for(i in 1:Ntip(treesub)){
    cc<-if(obj$xx[i]>0) 5 else -5
    th<-atan(obj$yy[i]/obj$xx[i])
    segments(obj$xx[i],obj$yy[i],
        obj$xx[i]+cc*cos(th),
        obj$yy[i]+cc*sin(th),lwd=4,
        col=cols[fac[treesub$tip.label[i]]])
}
add.color.bar(max(nodeHeights(treesub)),dMap$cols,
    prompt=FALSE,lwd=6,outline=FALSE,x=0.95*par()$usr[1],
    y=0.9*par()$usr[4],subtitle="",fsize=0.8)
h<-max(nodeHeights(treesub))
ii<-order(rowSums(pp^2),decreasing=TRUE)
for(i in ii){
    plotrix::floating.pie(obj$xx[i+Ntip(treesub)],
        obj$yy[i+Ntip(treesub)],
        radius=if(max(pp[i,])<=0.95) 0.03*h else 0.01*h,
        x=pp[i,],col=cols,border="transparent")
}
legend(x=0.98*par()$usr[1],y=0.9*par()$usr[4],
    levels(fac),pch=16,col=cols,
    pt.cex=2,bty="n",cex=0.8)
```

Unlike the 5-state trait, some ancestral states for this binary trait, especially within suboscines, are ambiguous; let's look at the nodes more closely.

- Extract values for pie charts at the nodes for the binary trait
```{r}
# Visualize the nodes in the tree
ggtree(treesub) + 
  geom_text2(aes(subset=!isTip, label=node), hjust=-.3) +
  geom_tiplab()

# Check posterior probabilities of ancestral states
ASRbinary<-as.data.frame(cbind(rownames(objstudy$ace), objstudy$ace))
ASRbinary[,1:3]<-as.numeric(unlist(ASRbinary[,1:3]))

## Nodes of interest
clade<-c("Aves", "Trochilidae", "Telluraves", "Australaves", "Eufalconimorphae", "Falconiformes", "Psittacopasseres", "Psittaciformes", "Passeriformes", "Eupasseres", "Tyranni", "Eurylaimides", "Tyrannides", "Tyrannida",  "Furnariida", "Passeri", "Passeri w/o (Menuroidea)")
node<-c(1114, 2210, 1118, 1119, 1120, 2179, 1121, 2126, 1122, 1123, 1658, 2094, 1659, 1756, 1660, 1124, 1125)

## Subset in correct order
ASRbinary<-ASRbinary %>% 
  slice(order(factor(V1, levels = node))) %>% 
  filter(V1 %in% node)
rownames(ASRbinary)<-clade; colnames(ASRbinary)<-c("Node number", "p (0)", "p (1)")
ASRbinary
```

- Same for the 5-state trait, just in case
```{r}
ASR5<-as.data.frame(cbind(rownames(objstudy1$ace), objstudy1$ace))
ASR5[,1:6]<-as.numeric(unlist(ASR5[,1:6]))

ASR5<-ASR5 %>% 
  slice(order(factor(V1, levels = node))) %>% 
  filter(V1 %in% node)
ASR5<-signif_df(ASR5)
colnames(ASR5)<-c("Node number", "p (0)", "p (1)", "p (2)", "p (3)", "p (4)")
rownames(ASR5)<-clade
ASR5
```

